---
title: "Variational Approximations in Survival Analysis"
bibliography: ref.bib
biblio-style: apa
output: 
  revealjs::revealjs_presentation:
    css: styles.css
    theme: black
    center: false
    transition: slide
    highlight: monochrome
    self_contained: true
    reveal_options:
      slideNumber: true
    includes:
      in_header: header.html
      after_body: doc_suffix.html
---

## dummy slide

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 5, cache.path = "cache/", 
                      message = FALSE, error = FALSE, warning = FALSE)
.par_use <- list(cex = 1.33, cex.lab = 1.2)
options(digits = 3, knitr.kable.NA = '')
```

<!--html_preserve-->
<script>
(function() {
  document.getElementById("dummy-slide").remove(); 
  
  var front_div = document.getElementsByTagName("section")[0];
  front_div.classList.add("front");
  front_div.classList.add("center");
  
  // conference/where this is at
  var where_at = document.createElement("p");
  var where_at_text = document.createElement("i");
  var node = document.createTextNode("SeRC Meeting 2021");
  where_at_text.appendChild(node);
  where_at.appendChild(where_at_text);
  where_at.style.margin = "0.1em";
  where_at.style.fontSize = "75%";
  front_div.appendChild(where_at);
  
  // add author 
  var credit_div = document.createElement('div');
  credit_div.innerHTML += "<div class='w-small'><p>Benjamin Christoffersen</p><p class='smallish'>KI, Department of Medical Epidemiology and Biostatistics, <a href='mailto:benjamin.christoffersen@ki.se'>benjamin.christoffersen@ki.se</a></p><p class='smallish'>KTH, Division of Robotics, Perception and Learning, <a href='mailto:benchr@kth.se'>benchr@kth.se</a></p></div>";
  credit_div.classList.add("authors");
  front_div.appendChild(credit_div);
})();
</script>
<!--end dummy slide-->

</section>

<section>
<section class="large-first center slide level2">
<h1>Project Overview</h1>
<!--/html_preserve-->

<div style="display: none;">
$$
\renewcommand\vec{\boldsymbol}
\def\bigO#1{\mathcal{O}(#1)}
\def\Cond#1#2{\left(#1\,\middle|\, #2\right)}
\def\mat#1{\boldsymbol{#1}}
\def\der{{\mathop{}\!\mathrm{d}}}
\def\argmax{\text{arg}\,\text{max}}
\def\Prob{\text{P}}
\def\Expec{\text{E}}
\def\logit{\text{logit}}
\def\diag{\text{diag}}
$$
</div>

## General Problem: Fitting Mixed Models

<div class = "w-small">
Have outcomes $\vec Y$ 
<p class = "smallish">
plausibly of heterogeneous types. E.g. a complication (yes/no), 
time-to-diagnosis (continuous or interval), or blood measurements
(continuous).</p>
</div>

<div class = "fragment w-small">
Relate the outcomes with an unobserved random effects $\vec U$ with density 
$h_{\vec\omega}(\vec u)$ and a conditional density 
$g_{\vec\omega}(\vec y \mid \vec u)$ 
<p class = "smallish">
depending model parameter $\vec\omega$.</p></div>

<div class = "fragment w-small">
Thus, the log likelihood is 

$$
l(\vec y,\vec\omega) = \log \int h_{\vec\omega}(\vec u)g_{\vec\omega}(\vec y \mid \vec u)d\vec u.
$$
<p class = "smallish">
Needed for model estimation. $g$ and $h$ are usually easy to evaluate pointwise 
but no closed form solution.</p>
</div>

## Issue

Current methods and implementations are often slow or too biased. 

<p class = "fragment">
Researchers restrict their models and questions because of these 
constraints.</p>

## Non-Variational Approximations Work
<div class = "w-small">
Study of approximation methods for a commonly used and broad class of mixed 
models.
<p class = "smallish">
Draft has been submitted.</p></div>

<div class = "w-small fragment">
Lead to imputation method for mixed data types. 
<p class = "smallish">
Article is accepted at the Asian Conference on Machine Learning. The
software has 3197 downloads and a draft is at @christoffersen2021asymptotically.</p>
</div>

## Non-Variational Approximations Work

<div class = "w-small ">
Lead to a fast and precise estimation method for common heritability models. 
<p class = "smallish">
Mature draft to be submitted. Started collaborations with Behrang Mahjani 
(ISMMS, New York), Benjamin Yip (CUHK; Hong Kong), and Sven Sandin 
(ISMMS, New York).
</p></div>

<div class = "fragment">
Data sets include: 

 - The Swedish National Patient Register.
 - The Swedish Multi-Generation Registry.

</div>

## Variational Approximations
Replace the log likelihood with a lower bound for some $\vec\theta\in\Theta$:

<div class = "w-small fragment">

$$
\begin{align*}
l(\vec y, \vec\omega)
  &\geq \tilde l(\vec y, \vec\omega, \vec\theta).
\end{align*}
$$
<p class = "smallish">
$\tilde l(\vec y, \vec\omega, \vec\theta)$ usually requires one-dimensional 
integration at worst.
</p></div>

<p class = "fragment">
A "by-product" is an approximation of the conditional distribution of the random 
effects given the data.</p>

<div class = "w-small fragment">
Typically have $n$ outcomes
<p class = "smallish">
and work with $l^{(i)}$, $h^{(i)}_{\vec\omega}$, 
$g^{(i)}_{\vec\omega}$, 
$q_{\vec\theta_i}^{(i)}$, $\Theta_i$, and 
$\tilde l^{(i)}$.</p></div>

## Variational Approximations (Cont.)

Approximate maximum likelihood

$$
\text{arg max}_{\vec\omega} \sum_{i = 1}^{n}l^{(i)}(\vec y_i, \vec\omega) 
  \approx \text{arg max}_{\vec\omega} 
  \sum_{i = 1}^{n} \max_{\vec\theta_i}\tilde l^{(i)}(\vec y_i, \vec\omega, \vec\theta_i).
$$

<div class = "fragment w-small">
Very fast pointwise evaluation but many more parameters!
<p class = "smallish">
10,000 to millions with register data sets.</p></div>

<div class = "w-small fragment">
Gradient descent or limited-memory BFGS are too slow.
<p class = "smallish">
We have developed special methods.</p></div>

## Comparison with Machine Learning

ML goal is often the performance of the left-hand side of 

<div class = "w-small">
$$
E_{q_{\hat{\vec\theta}_i}}(f_{\hat{\vec\omega}}(\vec U)) \approx 
  E_{p_{\hat{\vec\omega}}}(g_{\hat{\vec\omega}}(\vec U))  
$$

<p class = "smallish">
where $q$ is a distribution chosen for the variational approximation and $p$ 
is the true conditional distribution of the random effects. Other quantities 
using $q_{\hat{\vec\theta}_i}$ are also used. The precision of the 
approximation is not the focus.</p></div>

<div class = "fragment w-small">
In statistics, the interest is usually on a precise approximation of

$$
\text{arg max}_{\vec\omega} \sum_{i = 1}^{n}l^{(i)}(\vec y_i, \vec\omega) 
  \approx \text{arg max}_{\vec\omega} 
  \sum_{i = 1}^{n} \max_{\vec\theta_i}\tilde l^{(i)}(\vec y_i, \vec\omega, \vec\theta_i).
$$

<p class = "smallish">
Often requires a tight lower bound! $\vec\omega$ is also of much lower 
dimension.</p>
</div>

## Concrete Work

<div class = "w-small">
Developed a variational approximations for a class of mixed survival models. 
<p class = "smallish">
Draft have been submitted.</p></div>

<div class = "fragment w-small">
Working on joint survival and marker models. 
<p class = "smallish">
Seen up to two orders of magnitude faster estimation and only small bias in 
preliminary studies. Plan collaboration with Juan Jesus Carrero (KI) and 
Maya Alsheh Ali (KI).</p>
</div>

<div class = "fragment">
Data sets include:

 - Electronic health records. 
 - Images. 

</div>

## Concrete Work (cont.)

<div class = "w-small">
Implemented a library to optimize partially separable functions.
<p class = "smallish">
A header-only C++ library with an R interface and 5142 downloads.</p></div>

<!--html_preserve-->
</section>
<!-- need extra end tag before next section -->
</section>


<section>
<section  class="large-first center slide level2">
<h1>Optimizing Variational Approximations</h1>
<!--/html_preserve-->

## Example of Computation Times
<div class = "w-small">
Toy example with a mixed logistic regression with six random effects per 
$n = 1000$ observations.
<p class = "smallish">
E.g. an observation is a student taking multiple tests, or multiple patients 
at the same doctor. Random effects: person or doctor specific effects.
</p></div>

<p class = "fragment">
Sampled five data sets.
</p>

## Example of Computation Times

```{r run_time_ex}
tab <- read.table(text ="Mean Meadian
 Laplace         105.3204  73.671
 GVA               2.5618   2.423
 GVA4   0.8038   0.768
 GVA_LBFGS        24.6358  23.298")
rownames(tab) <- c("lme4", "psqn (our)", "psqn (our; 4 threads)", "LBFGS")
knitr::kable(tab)
```

<p class = "smallish">
Running times are in seconds. lme4: a Laplace 
approximation from the lme4 package, psqn: uses our psqn package, and 
LBFGS: uses limited-memory BFGS. From 
https://github.com/boennecd/psqn-va-ex.</p>

<p class = "fragment">
The Laplace approximation even has higher bias in this example!</p>

## Methods

1. Recursive method. 
2. Newton's method. 
3. Quasi-Newton using the partial separability.

## Recursive Method

Solve 
$\widehat{\vec\theta}_i(\vec\omega) = \text{arg max}_{\vec\theta_i}\tilde l^{(i)}(\vec y_i, \vec\omega, \vec\theta_i)$
for $i = 1,\dots,n$.

<div class = "fragment">
Take one step of Newton’s method to update $\vec\omega$ to find the maximum of

$$
\text{arg max}_{\vec\omega} 
  \sum_{i = 1}^n\tilde l^{(i)}\left(\vec y_i, \vec\omega, \widehat{\vec\theta}_i(\vec\omega)
  \right)
$$
</div>

<div class = "fragment w-small">
Repeat if not converged.
<p class = "smallish">
The Hessian in the second step can be efficiently 
computed.
Similar to the method suggested by @Ormerod12.</p>
</div>

## Newton's Method 
Use Newton's method to solve

$$
\text{arg max}_{\vec\omega,\vec\theta_1,\dots,\vec\theta_n} 
  \sum_{i = 1}^{n} \tilde l^{(i)}(\vec y_i, \vec\omega, \vec\theta_i).
$$

<div class = "fragment w-small">
Solve the Hessian using e.g. conjugate gradient.
<p class = "smallish">
The Hessian is very sparse and has a arrowhead matrix-like structure.
</p></div>

## Using the Partial Separability
<div class = "w-small">
Approximate the Hessian of each $\tilde l^{(i)}$ using BFGS.
<p class = "smallish">
Do not have to compute or implement the Hessian.</p>
</div>


<div data-fragment-index=1 class = "fragment">
Use a quasi-Newton method to solve 

$$
\text{arg max}_{\vec\omega,\vec\theta_1,\dots,\vec\theta_n} 
  \sum_{i = 1}^{n} \tilde l^{(i)}(\vec y_i, \vec\omega, \vec\theta_i) 
$$
</div>

<div class = "fragment w-small" data-fragment-index=1>
with the Hessian approximation. Solve with conjugate gradient. 
<p class = "smallish">
Implemented in the psqn package. Used in the example.</p>
</div>


## Remarks
The problems are commonly

 - embarrassing parallel. 
 - easy to make cache-friendly implementation for.
 
<div class = "fragment w-small">
Automatic differentiation can be used for the gradient or Hessian.
<p class = "smallish">
A fast implementation like @Hogan14 gives only a moderate overhead. We are 
using the very similar implementation from @savine2018modern. 
</p></div>

## Summary 

Variational approximation can be used to enable researchers to fit models that 
where not possible before...

<p class = "fragment">
... but this requires special methods for optimizing the lower bound.</p>

<!--html_preserve-->
</section>
<!-- need extra end tag before next section -->
</section>


<section>
<section class="center final">
<h1>Thank You!</h1>

<div class="w-small">
<p class="smallish">The presentation is at  
<a href="https://rpubs.com/boennecd/SeRC-meeting-2021">rpubs.com/boennecd/SeRC-meeting-2021</a>.</p>
<p class="smallish">The markdown is at  
<a href="https://github.com/boennecd/Talks">github.com/boennecd/Talks</a>.</p>
<p class="smallish">The psqn package is on CRAN and at 
<a href="https://github.com/boennecd/psqn">github.com/boennecd/psqn</a>.</p>
<p class="smallish">The imputation package, mdgc, is on CRAN and at 
<a href="https://github.com/boennecd/mdgc">github.com/boennecd/mdgc</a>.</p>
<p class="smallish">References are on the next slide.</p>
</div>

</section>
<!-- need extra end tag before next section -->
</section>

<section>
<h1>References</h1>

<!--/html_preserve-->
