---
title: "Computational Concerns with Mixed Models"
output: 
  revealjs::revealjs_presentation:
    css: styles.css
    theme: black
    center: false
    transition: slide
    highlight: monochrome
    self_contained: true
    reveal_options:
      slideNumber: true
    includes:
      in_header: header.html
      after_body: doc_suffix.html
---

## dummy slide

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 5, cache.path = "cache/", 
                      message = FALSE, error = FALSE, warning = FALSE)
.par_use <- list(cex = 1.33, cex.lab = 1.2)
options(digits = 3, knitr.kable.NA = '')
```

<!--html_preserve-->
<script>
(function() {
  document.getElementById("dummy-slide").remove(); 
  
  var front_div = document.getElementsByTagName("section")[0];
  front_div.classList.add("front");
  front_div.classList.add("center");
  
  // conference/where this is at
  var where_at = document.createElement("p");
  var where_at_text = document.createElement("i");
  var node = document.createTextNode("Group Meeting 2021");
  where_at_text.appendChild(node);
  where_at.appendChild(where_at_text);
  where_at.style.margin = "0.1em";
  where_at.style.fontSize = "75%";
  front_div.appendChild(where_at);
  
  // add author 
  var credit_div = document.createElement('div');
  credit_div.innerHTML += "<div class='w-small'><p>Benjamin Christoffersen</p><p class='smallish'>KI, Department of Medical Epidemiology and Biostatistics, <a href='mailto:benjamin.christoffersen@ki.se'>benjamin.christoffersen@ki.se</a></p><p class='smallish'>KTH, Division of Robotics, Perception and Learning, <a href='mailto:benchr@kth.se'>benchr@kth.se</a></p></div>";
  credit_div.classList.add("authors");
  front_div.appendChild(credit_div);
})();
</script>
<!--end dummy slide-->

</section>

<section>
<section class="large-first center slide level2">
<h1>Introduction</h1>
<!--/html_preserve-->

<div style="display: none;">
$$
\renewcommand\vec{\boldsymbol}
\def\bigO#1{\mathcal{O}(#1)}
\def\Cond#1#2{\left(#1\,\middle|\, #2\right)}
\def\mat#1{\boldsymbol{#1}}
\def\der{{\mathop{}\!\mathrm{d}}}
\def\argmax{\text{arg}\,\text{max}}
\def\Prob{\text{P}}
\def\Expec{\text{E}}
\def\logit{\text{logit}}
\def\diag{\text{diag}}
$$
</div>

## Running Example

Students attempt multiple tasks. We want to know: 

 1. How hard is each task?
 2. How much do the skill level differs between students?

## Running Example (Cont.)

Unobserved $\vec U_i$ is the skill level of student $i$. 

<div class = "fragment w-small">
Assume that each $\vec U_i$ has density $g(\cdot;\vec\theta)$.
<p class = "smallish">
Parts of $\vec\theta$ quantifies the difference in skill level.
</p></div>

<div class = "fragment">
Given skill level $\vec U_i$, the outcomes $\vec Y_i$ 
(whether each task is completed) has density 
$h(\cdot\mid\vec u;\vec\theta)$.
<p class = "smallish">
Parts of $\vec\theta$ quantifies the difficulty of each test.
</p></div>

<p class = "fragment">
Often care only about $\vec\theta$ and not $\vec U_i$.</p>

## Common Mixed Effect Model

The likelihood for individual $i$ is

$$
\exp f(\vec y_i, \vec u_i;\vec\theta) = g(\vec u_i;\vec\theta) 
  h(\vec y_i\mid\vec u_i;\vec\theta)
$$

<p class = "fragment">
$g(\vec u_i;\vec\theta)$ is similar to a prior in a Bayesian analysis.
</p>

<div class = "fragment">

The log marginal likelihood for individual $i$ is

$$
l(\vec\theta;\vec y_i) = \log \int \exp(f(\vec y_i, \vec u;\vec\theta))\der \vec u
$$

Needed to find the maximum likelihood estimator of $\vec\theta$.
</div>

## Running Example (Cont.)

Assume $\vec y_i\in\{0,1\}^{n}$ is whether the answer to each task is correct.

<div class = "fragment">

Use a mixed logistic regression

$$
\begin{align*}
f(\vec y_i,u_i;\vec\theta) &=
  \log\phi(u_i;\sigma^2) 
   \\ 
&\hspace{25pt}
   +\sum_{j = 1}^{n}
   \Big((\eta_j + u_i)y_{ij}
   -\log(1 + \exp(\eta_j + u_i))\Big)
\end{align*}
$$

$\eta_j$: difficulty of task $j$. 

$\sigma^2$: difference between students' skill level.
</div>

## Concrete example
<div class = "small">
Three tasks with $(\eta_1,\eta_2,\eta_3) = (-2, 0, 2)$. 
<p class = "smallish">
Assumed know but almost never known in practice and have to be estimated.</p>
</div>

<p class = "fragment">
Need an approximation for the log marginal likelihood
$l(\vec\theta;\vec y_i) = l(\sigma^2;\vec y_i)$</p>

<p class = "fragment">
We will suppress the dependence on $\vec y_i$ and $i$ and write 
e.g. $f(u;\vec\theta)$.</p>

<p class = "fragment">
Use one dimensional $U\in\mathbb R$ for simplicity.</p>


<!--html_preserve-->
</section>
<!-- need extra end tag before next section -->
</section>


<section>
<section class="large-first center slide level2">
<h1>Laplace Approximation</h1>
<!--/html_preserve-->

## Laplace Approximation

<div class = "w-small">
$$
\begin{align*}
l(\vec\theta) &= \log\int\exp(f(\vec u;\vec\theta))\der \vec u \\
              &\approx \frac K2\log 2\pi - 
              \frac 12\log
              \lVert -f''_{\vec u\vec u}(\vec u_0(\vec\theta);\vec\theta)\rVert
              + f(\vec u_0(\vec\theta);\vec\theta) \\
              &= \tilde L(\vec\theta)
\end{align*}
$$

<p class = "smallish">
where $\vec u \in \mathbb R^K$.</p>
</div>

<div class = "fragment">

$$\vec u_0(\vec\theta) = \text{arg max}_\vec u f(\vec u;\vec\theta)$$

<div class = "w-small">
and $f''_{\vec u\vec u}(\vec u_0;\vec\theta)$ is negative definite. 
<p class = "smallish">
Needed for a maximum and in the proof of the approximation.</p>
</div></div>

## Running Example

$$
\begin{align*}
f'_u(\vec y,u;\vec\theta) &= -\frac u{\sigma^2} + 
  \sum_{i = 1}^{n_i}\left( y_i - 
  \frac{\exp(\eta_i + u)}{1 + \exp(\eta_i + u)}\right) \\
f''_{uu}(\vec y,u;\vec\theta) &= -\frac 1{\sigma^2} - \sum_{i = 1}^{n_i}
  \frac{\exp(\eta_i + u)}{(1 + \exp(\eta_i + u))^2} \\
\tilde L(\vec\theta) &= \frac 12\log 2\pi - 
              \frac 12\log
              \lVert -f''_{uu}(u_0(\vec\theta);\vec\theta)\rVert
              + f(u_0(\vec\theta);\vec\theta)
\end{align*}
$$

</section>
<section class="center-horiz" data-transition="slide-in slide-out">
<h2>Example: Laplace Approximation</h2>

```{r setup_n_laplace, fig.keep="last"}
# the code is not meant to be fast and could be improved in many ways!

# the complete log likelihood, f
complete_ll <- Vectorize(\(u, eta, sig, y){
  eta <- eta + u
  log_exp1 <- ifelse(eta > 30, eta, log(1 + exp(eta)))
  dnorm(u, sd = sig, log = TRUE) + sum(eta * y - log_exp1)
}, "u")

# the marginal likelihood
marg_ll <- \(eta, sig, y)
  integrate(\(u) exp(complete_ll(u, eta = eta, sig = sig, y = y)), 
            -Inf, Inf, rel.tol = .Machine$double.eps^(1/2))$value

# parameter setting
eta <- c(-2, 0, 2)
sig <- 1
combs <- as.matrix(expand.grid(c(0, 1), c(0, 1), c(0, 1)))

true_probs <- apply(combs, 1, marg_ll, eta = eta, sig = sig)

# get points for the true curve
sig_points <- seq(.1, 2, length.out = 20)
true_vals <- sapply(sig_points, \(sig){
  probs <- apply(combs, 1, marg_ll, eta = eta, sig = sig)
  sum(true_probs * log(probs))
})

plot_true_curve <- \(){
  par(mar = c(5, 5, 1, 1))
  plot(sig_points, true_vals, pch = 16, xlab = expression(sigma), bty = "l",
       ylab = expression(paste("Log Marginal Likelihood, ", l(theta))))
  lines(spline(sig_points, true_vals))
  abline(v = sig, lty = 2)
  grid()
}
plot_true_curve()

# add the Laplace approximation
laplace <- \(sig){
  # apply the Laplace approximation for each y
  out <- apply(combs, 1, \(y){
    u_max <- optimize(\(u) -complete_ll(u, eta = eta, sig = sig, y = y), 
                      c(-10, 10), tol = .Machine$double.eps)
    
    f <- -u_max$objective
    eta <- eta + u_max$minimum
    fpp <- -1/sig^2 - sum(exp(eta) / (1 + exp(eta))^2)
    
    log(2 * pi) / 2 - log(-fpp) / 2 + f
  })
  
  sum(true_probs * out)
}

add_laplace <- \(){
  laplace_points <- sapply(sig_points, \(sig) laplace(sig))
  points(sig_points, laplace_points, pch = 17)
  lines(spline(sig_points, laplace_points), lty = 2)
}
add_laplace()
```

<div style = "text-align:left;">
<p class = "smallish">
The continuous line is true curve and the dashed line is the Laplace approximation.
</p></div>

## Remarks

$$
\begin{align*}
l(\vec\theta) \approx \frac K2\log 2\pi - 
              \frac 12\log
              \lVert -f''_{\vec u\vec u}(\vec u_0(\vec\theta);\vec\theta)\rVert
              + f(\vec u_0(\vec\theta);\vec\theta)
\end{align*}
$$

Finding $\vec u_0(\vec\theta) = \text{arg max}_\vec u f(\vec u;\vec\theta)$
is often fast.

<p class = "fragment">
The Laplace approximation is fast but may be biased.</p>

<div class = "fragment w-small">
Has nice asymptotics in many cases.
<p class = "smallish">
In the running example when there are many tasks done by each student.</p>
</div>

## Laplace Approximation (Cont.)

$$
\begin{multline*}
\tilde L'_{\vec\theta}(\vec\theta) =
  f'_{\vec\theta}(\vec u_0(\vec\theta);\vec\theta)
  +\frac 12\left(
  \nabla_{\vec u}\log
  \lVert -f''_{\vec u\vec u}(\vec u_0(\vec\theta);\vec\theta)\rVert
  \right) \\
  f''_{\vec u\vec u}(\vec u_0(\vec\theta);\vec\theta)^{-1}
  f''_{\vec u\vec\theta}(\vec u_0(\vec\theta);\vec\theta)
\end{multline*}
$$

Point: requires 3rd order derivatives for the gradient.

<div class = "fragment w-small">
Automatic differentiation implementations
with higher order derivatives often have a big overhead.
<p class = "smallish">
A general approach has been implemented in the `TMB` package.</p>
</div>


<!--html_preserve-->
</section>
<!-- need extra end tag before next section -->
</section>


<section>
<section class="large-first center slide level2">
<h1>Gaussian Quadrature</h1>
<!--/html_preserve-->

## Gaussian Quadrature

$$
\int \omega(u)g(u)\der u \approx \sum_{i = 1}^m w_ig(u_i)
$$

with nodes $u_1,\dots,u_m$ and weights $w_1,\dots,w_m$. 

<p class = "fragment">
The approximation is exact for polynomials of degree $2m - 1$ or less.</p>

<p class = "fragment">
Gauss–Hermite quadrature is often used where $\omega(x) = \exp(-x^2)$.</p>

## Application

$$
\begin{align*}
\int\exp f(\vec y,u;\vec\theta) \der u &=
  \frac 1{\sqrt{2\pi}\sigma}\int
  \exp\left(
  -\frac{u^2}{2\sigma^2}\right)
  h(\vec y\mid u;\vec\theta)\der u \\
&=  \frac 1{\sqrt{\pi}}\int
 \underbrace{\exp\left(-u^2\right)}_{\omega(u)}
  h(\vec y\mid \sqrt 2 \sigma u;\vec\theta)\der u
\end{align*}
$$

</section>
<section class="center-horiz" data-transition="slide-in slide-out">
<h2>Application (Cont.)</h2>

```{r ghq, fig.show = 'animate', interval=2}
library(gaussquad)
rules <- hermite.h.quadrature.rules(8)

use_gh_rule <- \(rule, sig){
  out <- apply(combs, 1, \(y){
    out <- sapply(rule$x, \(u){
      eta <- eta + sqrt(2) * sig *  u
      log_exp1 <- ifelse(eta > 30, eta, log(1 + exp(eta)))
      exp(sum(eta * y - log_exp1)) / sqrt(pi)
    })
    log(sum(out * rule$w))
  })
  
  sum(true_probs * out)
}

for(r in rules){
  plot_true_curve()
  add_laplace()
  
  # add GHQ points
  vs <- sapply(sig_points, \(s) use_gh_rule(r, sig = s))
  points(sig_points, vs, pch = 18)
  lines(spline(sig_points, vs), lty = 3)
  text(.4, -1.59, labels = sprintf("%2d nodes", NROW(r$x)), cex = 2)
}
```

<div style = "text-align:left;">
<p class = "smallish">
The continuous line is true curve, the dashed line is the Laplace approximation,
and the dotted line is the Gauss-Hermite quadrature.
</p></div>

</section>
<section class="center-horiz" data-transition="slide-in fade-out">
<h2>Adaptive Version</h2>

```{r 2pre_adaptive_illustration}
local({
  y <- combs[1, ]
  
  par(mar = c(5, 5, 1, 1))
  plot(\(u) exp(complete_ll(u, eta = eta, sig = sig, y = y)), 
       xlim = c(-4, 4), 
       ylab = expression(paste(exp(f(u, theta)), " or density")), 
       xlab = expression(u), 
       bty = "l", yaxs = "i", xaxs = "i", yaxt = "n")
  grid()
  
  pts <- seq(-4, 4, length.out = 100)
  dnrms <- dnorm(pts)
  lines(pts, dnrms * par("usr")[4] * .98 / max(dnrms), lty = 2)  
  
  rug(rules[[length(rules)]]$x, ticksize = .1, lwd = 2)
})
```

<div style = "text-align:left;">
<p class = "smallish">
The integrand (continuous) and $g(\vec u_i;\vec\theta)$ (dashed). 
The ticks on the first axis are the quadrature nodes, $u_i$.
</p></div>

</section>
<section class="center-horiz" data-transition="fade-in slide-out">
<h2>Adaptive Version</h2>

```{r 3pre_adaptive_illustration}
local({
  y <- combs[1, ]
  
  par(mar = c(5, 5, 1, 1))
  plot(\(u) exp(complete_ll(u, eta = eta, sig = sig, y = y)), 
       xlim = c(-4, 4), 
       ylab = expression(paste(exp(f(u, theta)), " or density")), 
       xlab = expression(u), 
       bty = "l", yaxs = "i", xaxs = "i", yaxt = "n")
  grid()
  
  u_max <- optimize(\(u) -complete_ll(u, eta = eta, sig = sig, y = y), 
                      c(-10, 10), tol = .Machine$double.eps)
    
  eta_p <- eta + u_max$minimum
  fpp <- -1/sig^2 - sum(exp(eta_p) / (1 + exp(eta_p))^2)
  scale <- sqrt(2 * (1/-fpp))
  x_use <- scale * rules[[length(rules)]]$x + u_max$minimum
  
  pts <- seq(-4, 4, length.out = 100)
  dnrms <- dnorm(pts, mean = u_max$minimum, sd = scale / sqrt(2))
  lines(pts, dnrms * par("usr")[4] * .98 / max(dnrms), lty = 2)  
  
  rug(x_use, ticksize = .1, lwd = 2)
})
```

<div style = "text-align:left;">
<p class = "smallish">
The integrand (continuous) and the shifted and re-scaled weight (dashed). 
The ticks on the first axis are the quadrature nodes, $u_i$.
</p></div>

## Adaptive Version (Cont.)

$$
\begin{align*}
\int
  \exp\left(f\left(u;\vec\theta\right)\right)\der u 
&= \sqrt{2\hat\sigma^2(\vec\theta)}\int
    \exp(-u^2)\\
&\hspace{25pt}
    \cdot \exp\left(u^2 + f\left(
      \sqrt{2\hat\sigma^2(\vec\theta)} u + 
      u_0(\vec\theta);\vec\theta\right)\right) \der u \\
u_0(\vec\theta) &= \text{arg max}_u f\left(u;\vec\theta\right) \\
\hat\sigma^2(\vec\theta) &= 
  (-f''_{uu}(u_0(\vec\theta);\vec\theta))^{-1}
\end{align*}
$$

</section>
<section class="center-horiz" data-transition="slide-in slide-out">
<h2>Application of the Adaptive Version</h2>

```{r aghq, fig.show = 'animate', interval=2}
use_agh_rule <- \(rule, sig){
  out <- apply(combs, 1, \(y){
    u_max <- optimize(\(u) -complete_ll(u, eta = eta, sig = sig, y = y), 
                      c(-10, 10), tol = .Machine$double.eps)
    
    eta_p <- eta + u_max$minimum
    fpp <- -1/sig^2 - sum(exp(eta_p) / (1 + exp(eta_p))^2)
    scale <- sqrt(2 * (1/-fpp))
    x_use <- scale * rule$x + u_max$minimum
    
    out <- sapply(x_use, complete_ll, eta = eta, sig = sig, y = y)
    log(sum(exp(out + rule$x^2) * rule$w * scale))
  })
  
  sum(true_probs * out)
}

for(r in rules){
  plot_true_curve()
  add_laplace()
  
  # GHQ points
  vs <- sapply(sig_points, \(s) use_gh_rule(r, sig = s))
  points(sig_points, vs, pch = 18)
  lines(spline(sig_points, vs), lty = 3)
  
  # AGHQ points
  vs <- sapply(sig_points, \(s) use_agh_rule(r, sig = s))
  points(sig_points, vs, pch = 19, col = "darkblue")
  lines(spline(sig_points, vs), lty = 3, col = "darkblue")
  
  text(.4, -1.59, labels = sprintf("%2d nodes", NROW(r$x)), cex = 2)
}
```

<div style = "text-align:left;">
<p class = "smallish">
The continuous line is true curve, the dashed line is the Laplace approximation,
and the dotted (green) line is the (adapted) Gauss-Hermite quadrature.
</p></div>

## Remarks

<div class = "w-small">
Often works very well with common models and few random effects.
<p class = "smallish">
If $\vec U\in\mathbb R^K$ then say $K \leq 5$.</p>
</div>

<div class = "w-small fragment">
The computational complexity is $\bigO{m^K}$.
<p class = "smallish">
$m$ is the number of quadrature nodes.</p>
</div>

<div class = "fragment w-small">
But many problems do not have many random effects.
<p class = "smallish">
$K$ is small.</p>
</div>

<!--html_preserve-->
</section>
<!-- need extra end tag before next section -->
</section>


<section>
<section class="large-first center slide level2">
<h1>Monte Carlo Approximation</h1>
<!--/html_preserve-->

## Simple Monte Carlo

$$
\int g(\vec u;\vec\theta) 
  h(\vec y\mid\vec u;\vec\theta) \der\vec u
$$

Draw from $g$, evaluate $h$, and take the average.

</section>
<section class="center-horiz" data-transition="slide-in slide-out">
<h2>Application of Monte Carlo</h2>

```{r MC_estimator, fig.show='animate', cache = 1, interval=2}
mc_estimate <- \(n_sample, sig, n_repeat = 25L)
  replicate(n_repeat, {
    out <- apply(combs, 1, \(y){
      out <- sapply(rnorm(n_sample, sd = sig), \(u){
        eta <- eta + u
        log_exp1 <- ifelse(eta > 30, eta, log(1 + exp(eta)))
        sum(eta * y - log_exp1)
      })
      mean(exp(out))
    })
    
    sum(log(out) * true_probs)
  })

set.seed(1)
n_sample_use <- 20 * 2^(0:5)
for(n_sample in n_sample_use){
  plot_true_curve()
  add_laplace()
  
  # add MC estimates
  mc_res <- sapply(sig_points, mc_estimate, n_sample = n_sample)
  matplot(sig_points, t(mc_res), pch = 19, add = TRUE, col = gray(0, .2))
  lines(smooth.spline (rep(sig_points, each = NROW(mc_res)), mc_res), lty = 3)
  
  text(.4, -1.59, labels = sprintf("%3d samples", n_sample), cex = 2)
}
```

<div style = "text-align:left;">
<p class = "smallish">
The continuous line is true curve, the dashed line is the Laplace approximation,
and the dots are Monte Carlo estimators. The dotted line is an average 
estimate.
</p></div>

## Importance Sampling

$$
\begin{align*}
\int
  \exp\left(f\left(u;\vec\theta\right)\right)\der u 
&= \int
  \phi\left(u;u_0(\vec\theta),\hat\sigma^2(\vec\theta)\right)
  \frac{\exp\left(f\left(
      u;\vec\theta\right)\right)}
      {\phi\left(u;u_0(\vec\theta),\hat\sigma^2(\vec\theta)\right)} \der u \\
u_0(\vec\theta) &= \text{arg max}_u f\left(u;\vec\theta\right) \\
\hat\sigma^2(\vec\theta) &= 
  (-f''_{uu}(u_0(\vec\theta);\vec\theta))^{-1}
\end{align*}
$$

Like the re-scaling and shifting with quadrature.

</section>
<section class="center-horiz" data-transition="slide-in slide-out">
<h2>Application of Importance Sampling</h2>

```{r import_estimator, fig.show='animate', cache = 1, interval=2}
import_estimate <- \(n_sample, sig, n_repeat = 25L)
  replicate(n_repeat, {
    out <- apply(combs, 1, \(y){
      u_max <- optimize(\(u) -complete_ll(u, eta = eta, sig = sig, y = y), 
                      c(-10, 10), tol = .Machine$double.eps)
    
      eta_p <- eta + u_max$minimum
      fpp <- -1/sig^2 - sum(exp(eta_p) / (1 + exp(eta_p))^2)
      scale <- sqrt(1/-fpp)
      
      out <- sapply(
        rnorm(n_sample, u_max$minimum, scale), 
        \(u) complete_ll(u, eta = eta, sig = sig, y = y) - 
          dnorm(u, u_max$minimum, scale, log = TRUE))
      mean(exp(out))
    })
    
    sum(log(out) * true_probs)
  })

set.seed(1)
for(n_sample in n_sample_use){
  plot_true_curve()
  add_laplace()
  
  # add importance sampler estimates
  imp_mc_res <- sapply(sig_points, import_estimate, n_sample = n_sample)
  matplot(sig_points, t(imp_mc_res), pch = 19, add = TRUE, col = gray(0, .2))
  lines(smooth.spline (rep(sig_points, each = NROW(imp_mc_res)), imp_mc_res), 
        lty = 3)
  
  text(.4, -1.59, labels = sprintf("%3d samples", n_sample), cex = 2)
}
```

<div style = "text-align:left;">
<p class = "smallish">
The continuous line is true curve, the dashed line is the Laplace approximation,
and the dots are importance sampling estimators. The dotted line is an average 
estimate.
</p></div>

</section>
<section class="center-horiz" data-transition="slide-in slide-out">
<h2>Error Term</h2>

```{r mc_show_error}
set.seed(1)
mc_res <- sapply(n_sample_use, mc_estimate, sig = sig)
imp_mc_res <- sapply(n_sample_use, import_estimate, sig = sig)
true_val_use <- true_vals[abs(sig_points - sig) < 1e-4]
mc_err <- abs(t(exp(mc_res)) - exp(true_val_use))
imp_mc_err <- abs(t(exp(imp_mc_res)) - exp(true_val_use))

par(mar = c(5, 5, 1, 1))
matplot(n_sample_use * .98, mc_err, col = "black", pch = 16, bty = "l",
        ylim = range(mc_err, imp_mc_err), xlab = "Number of smaples",
        log = "xy", ylab = "Marginal Likelihood error")
matplot(n_sample_use * 1.02, imp_mc_err, col = "darkblue", pch = 17, add = TRUE)

f_mc <- lm(c(log(mc_err)) ~ rep(log(n_sample_use), NCOL(mc_err)))
f_imp <- lm(c(log(imp_mc_err)) ~ rep(log(n_sample_use), NCOL(mc_err)))

lines(n_sample_use, exp(coef(f_mc)[1] + coef(f_mc)[2] * log(n_sample_use)))
lines(n_sample_use, exp(coef(f_imp)[1] + coef(f_imp)[2] * log(n_sample_use)),
      col = "darkblue")
```

<div style = "text-align:left;">
<p class = "smallish">
The error of the estimated marginal likelihood versus the number of samples. 
The Monte Carlo estimates are black and the importance sampling estimates
are blue. Lines are from log-log regressions.
</p></div>

## Remarks

Both estimates are $\bigO{m^{-1/2}}$ where $m$ is the number of samples
but the constant may differ greatly.

<p class = "fragment">
Very efficient importance samplers exists for some models.</p>

<div class = "fragment w-small">
Various variance reduction methods exists.
<p class = "smallish">
E.g. antithetic variates and control variates. Effects only the constant.
</div>

<div class = "fragment w-small">
Biased estimate of the log marginal likelihood
<p class = "smallish">
in finite samples. Usually not an issue.</p></div>

<p class = "fragment">
Markov chain Monte Carlo and related methods are for higher dimensional 
problems.</p>

## Quasi-Monte Carlo

Select numbers that are not random but a sequence with a low-discrepancy.

<p class = "fragment">
Estimates are $\bigO{m^{-1 + \epsilon}}$ where $m$ is the length of the 
sequence.
</p>

</section>
<section class="center-horiz" data-transition="slide-in slide-out">
<h2>Quasi-Monte Carlo</h2>

```{r QMC, fig.show='animate', cache = 1}
library(randtoolbox)
QMC_estimate <- \(seq_len, sig){
  rng_dim <- NROW(combs)
  rngs <- halton(seq_len)
    
  out <- sapply(1:rng_dim, \(i){
    y <- combs[i, ]
    u_max <- optimize(\(u) -complete_ll(u, eta = eta, sig = sig, y = y),
                      c(-10, 10), tol = .Machine$double.eps)

    eta_p <- eta + u_max$minimum
    fpp <- -1/sig^2 - sum(exp(eta_p) / (1 + exp(eta_p))^2)
    scale <- sqrt(1/-fpp)
    
    out <- sapply(
      qnorm(rngs, u_max$minimum, scale), 
      \(u) complete_ll(u, eta = eta, sig = sig, y = y) - 
        dnorm(u, u_max$minimum, scale, log = TRUE))
    mean(exp(out))
  })
  
  sum(log(out) * true_probs)
}

for(seq_len in n_sample_use){
  plot_true_curve()
  add_laplace()
  
  # add the QMC estimates
  qmc_res <- sapply(sig_points, QMC_estimate, seq_len = seq_len)
  points(sig_points, qmc_res, col = gray(0, .2))
  lines(spline(sig_points, qmc_res), lty = 3)
  
  text(.4, -1.59, labels = sprintf("%3d samples", seq_len), cex = 2)
}
```

<div style = "text-align:left;">
<p class = "smallish">
The continuous line is true curve, the dashed line is the Laplace approximation,
and the dotted line is the quasi-Monte Carlo estimates.
</p></div>

</section>
<section class="center-horiz" data-transition="slide-in slide-out">
<h2>Error Term: with Quasi-Monte Carlo</h2>

```{r qmc_show_error}
qmc_res <- sapply(n_sample_use, QMC_estimate, sig = sig)
qmc_err <- abs(t(exp(qmc_res)) - exp(true_val_use))

par(mar = c(5, 5, 1, 1))
matplot(n_sample_use * .98, mc_err, col = "black", pch = 16, bty = "l",
        ylim = range(mc_err, imp_mc_err), xlab = "Number of smaples",
        log = "xy", ylab = "Marginal Likelihood error")
matplot(n_sample_use * 1.02, imp_mc_err, col = "darkblue", pch = 17, add = TRUE)
points(n_sample_use * 1.04, qmc_err, pch = 18)

f_qmc <- lm(c(log(qmc_err)) ~ log(n_sample_use))

lines(n_sample_use, exp(coef(f_mc)[1] + coef(f_mc)[2] * log(n_sample_use)))
lines(n_sample_use, exp(coef(f_imp)[1] + coef(f_imp)[2] * log(n_sample_use)),
      col = "darkblue")
lines(n_sample_use, exp(coef(f_qmc)[1] + coef(f_qmc)[2] * log(n_sample_use)),
      lty = 2)
```

<div style = "text-align:left;">
<p class = "smallish">
The error of the estimated marginal likelihood versus the number of samples. 
The Monte Carlo estimates are black and the importance sampling estimates
are blue. Lines are from log-log regressions. A line is also added for the 
quasi-Monte Carlo estimate.
</p></div>


<!--html_preserve-->
</section>
<!-- need extra end tag before next section -->
</section>


<section>
<section class="large-first center slide level2">
<h1>Variational Approximations</h1>
<!--/html_preserve-->

## Variational Approximations

Pick some variational distribution with density $q(\cdot;\vec\omega)$ for 
some $\vec\omega\in\Omega$. Then:

$$
\begin{align*}
l(\vec\theta) &= 
  \log \int q(\vec u;\vec\omega)
    \log \left(\frac{p(\vec y, \vec u;\vec\theta)\big/ q(\vec u;\vec\omega)}
    {p(\vec u\mid\vec y)\big/ q(\vec u;\vec\omega)}\right) \der \vec u\\
  &= \log \int q(\vec u;\vec\omega)
    \left(
      \log \left(\frac{p(\vec y, \vec u;\vec\theta)}
      {q(\vec u;\vec\omega)}\right) +
      \log \left(\frac{q(\vec u;\vec\omega)}
      {p(\vec u\mid\vec y;\vec\theta)}\right)
    \right) \der \vec u \\
  &\geq \log \int q(\vec u;\vec\omega)
    \log \left(\frac{p(\vec y, \vec u;\vec\theta)}
    {q(\vec u;\vec\omega)}\right) \der \vec u 
  = \tilde l(\vec\theta, \vec\omega)
\end{align*}
$$

## Variational Approximations (Cont.)

$$
l(\vec\theta) \geq \text{arg max}_{\vec\omega} \tilde l(\vec\theta, \vec\omega)
$$

Maybe very tight (almost equal).


## Running Example

We need to evaluate:

The entropy: 
$-E_{q(\cdot;\vec\omega)}(\log(q(U;\vec\omega)))$.

The first two moments:
$E_{q(\cdot;\vec\omega)}(U^2)\big/2\sigma^2$ and 
$E_{q(\cdot;\vec\omega)}(U)y_i$.

The following expectation:
$E_{q(\cdot;\vec\omega)}(\log(1 + \exp(\eta_i + U)))$.

</section>
<section class="center-horiz" data-transition="slide-in slide-out">
<h2>Gaussian Variational Approximation</h2>

```{r GVA, cache = 1, fig.keep="last"}
gva_lb <- \(sig){
  out <- apply(combs, 1, \(y){
    # the lower bound function
    lb <- \(par){
      mu <- par[1]
      lambda <- exp(par[2])
      sqrt_lambda <- sqrt(lambda)
      
      # the entropy term
      out <- (log(2 * pi * lambda) + 1) / 2
      
      # the terms from the other random effect density
      out <- out - log(2 * pi * sig^2) / 2 - (lambda + mu^2) / (2 * sig^2)
      
      # the terms from the outcomes
      log_partition_terms <- sapply(eta, \(e){
        integrate(Vectorize(\(u){
          e <- e + u
          e <- if(e > 30) e else log(1 + exp(e))
          e * dnorm(u, mu, sqrt_lambda)
        }), -Inf, Inf, rel.tol = .Machine$double.eps^(1/2))$value
      })
      eta_use <- eta + mu
      
      out + sum(eta_use * y - log_partition_terms)
    }
    
    va_pars <- optim(c(0, 0), \(x) -lb(x))
    -va_pars$value
  })
  
  sum(out * true_probs)
}

# create plot with the GVA curve
plot_true_curve()
add_laplace()
gva_points <- sapply(sig_points, gva_lb)

add_gva <- \(){
  points(sig_points, gva_points, pch = 18)
  lines(spline(sig_points, gva_points), lty = 3)
}
add_gva()
```

<div style = "text-align:left;">
<p class = "smallish">
The continuous line is true curve, the dashed line is the Laplace approximation,
and the dotted line is the lower bound from the Gaussian variational 
approximation.
</p></div>

</section>
<section class="center-horiz" data-transition="slide-in slide-out">
<h2>Skew-normal VA</h2>

```{r skeq_normal, fig.keep='last', cache = 1}
snva_lb <- \(sig){
  out <- apply(combs, 1, \(y){
    # the lower bound function
    lb <- \(par){
      mu <- par[1]
      lambda <- exp(par[2])
      skew <- par[3]
      sqrt_lambda <- sqrt(lambda)
      delta <- lambda * skew / sqrt(1 + skew^2 * lambda)
      
      # the entropy term
      entropy_sd <- sqrt(skew^2 * lambda)
      out <- (log(2 * pi * lambda) + 1) / 2 - log(2) - 
        2 * integrate(
          \(u) dnorm(u, sd = entropy_sd) * pnorm(u) * pnorm(u, log = TRUE),
          -Inf, Inf, rel.tol = .Machine$double.eps^(1/2))$value
      
      # the terms from the other random effect density
      mea <- mu + sqrt(2 / pi) * delta
      va <- lambda - 2 / pi * delta^2
      out <- out - log(2 * pi * sig^2) / 2 - (va + mea^2) / (2 * sig^2)
      
      # the terms from the outcomes
      log_partition_terms <- sapply(eta, \(e){
        integrate(Vectorize(\(u){
          e <- e + u
          e <- if(e > 30) e else log(1 + exp(e))
          e * 2 * dnorm(u, mu, sqrt_lambda) * pnorm(skew * (u - mu))
        }), -Inf, Inf, rel.tol = .Machine$double.eps^(1/2))$value
      })
      eta_use <- eta + mea
      
      out + sum(eta_use * y - log_partition_terms)
    }
    
    va_pars <- optim(c(0, 0, .1), \(x) -lb(x))
    if(va_pars$convergence != 0)
      va_pars <- optim(c(0, 0, .1), \(x) -lb(x), method = "BFGS")
    -va_pars$value
  })
  sum(out * true_probs)
}

# create plot with the GVA curve
plot_true_curve()
add_laplace()
add_gva()

snva_points <- sapply(sig_points, snva_lb)

add_snva <- \(){
  points(sig_points, snva_points, pch = 18, col = "darkblue")
  lines(spline(sig_points, snva_points), lty = 3, col =  "darkblue")
}
add_snva()
```

<div style = "text-align:left;">
<p class = "smallish">
The continuous line is true curve, the dashed line is the Laplace approximation,
and the (blue) dotted line is the lower bound from the 
Gaussian (skew normal) variational approximation.
</p></div>

## Remarks

Works quite well in this case but this is not universally true. 

<div class = "w-small fragment">
The computational complexity of the lower bound is often $\bigO{K^2}$ or 
$\bigO{K^3}$ at worst
<p class = "smallish">
where $\vec U\in\mathbb R^K$.</p>
</div>

<p class = "fragment"> 
A more flexible variational distribution can be used if the lower bounds is 
not tight enough.</p>

<div class = "fragment w-small"> 
Usually, never has no more than intractable one-dimensional integrals
<p class = "smallish">
where quadrature can be used.
</p></div>


<!--html_preserve-->
</section>
<!-- need extra end tag before next section -->
</section>



<section>
<section class="large-first center slide level2">
<h1>Summary</h1>
<!--/html_preserve-->

## Summary
An intractable likelihood is common with mixed models. 

<p class = "fragment">
The Laplace approximations is fast but may be biased.</p>

<p class = "fragment">
Gaussian quadrature works well for many low dimensional problems.</p>

<p class = "fragment">
Monte Carlo approximation are widely applicable but naive applications can have 
high variance.</p>

<p class = "fragment">
Variational approximation are fast but may be biased.</p>


<!--html_preserve-->
</section>
<!-- need extra end tag before next section -->
</section>


<section>
<section class="center final">
<h1>Thank You!</h1>

<div class="w-small">
<p class="smallish">The presentation is at
<a href="https://rpubs.com/boennecd/Comp-Mixed-Models">rpubs.com/boennecd/Comp-Mixed-Models</a>.</p>
<p class="smallish">The markdown is at
<a href="https://github.com/boennecd/Talks">github.com/boennecd/Talks</a>.</p>
</div>
<!--/html_preserve-->
